\chapter*{Acknowledgments}
\pdfbookmark{Acknowledgments}{acknowledgments} % Sets a PDF bookmark for the acknowledgment
\label{chap:acknowledgments}

There are many people without whose advice, help and support I would never have
been able to finish this thesis.

I can never overstate my gratitude towards my thesis advisor Andrei Beloborodov,
who has been ever supportive and patient, even in the darkest of times. An
ancient Chinese proverb says ``A teacher for a day is a father for a lifetime.''
I look up to Andrei as a fatherly figure, both academically and personally, who
taught me not only how to think about physics problems, but how to cope with
problems in life as well.

I would like to express my gratitude towards Professor Anatoly Spitkovksy, from
whom I learned about numerical methods and their potential pitfalls. He also
helped me tremendously in the process of postdoc application. Thanks also to
Professor Lorenzo Sironi, who helped me a lot with my job applications too, and
kindly agreed to serve on my thesis committee.

I would like to thank the colleagues whom I shared many inspiring discussions
with. I want to thank Rui Hu, who invested so much time working together with me
on this challenging coding project. Thank you to Kyle Parfrey, Romain Hascoët,
Indrek Vurm, Christoffer Lundman, Jonathan Zrake, and Xinyu Li; I greatly
enjoyed my discussions with you. Thank you to Mickey McDonald and Professor
Jeremy Dodd, it was a great pleasure working with you as a preceptor. I would
also like to thank my fellow grad students Zhe Liu, Xiao Xiao, Lian Liu, Daiqian
Zhang, Shuo Zhang, Xinyuan Ai, Lei Zhou, Jonghee Kang, Carlos Forsythe, Sathish
Thiyagarajan, for the fellowship we shared on our common paths towards the PhD.

There are many friends whose company I cherish over this long journey. To
Richard Xie, thank you for always being there for me. To Iris Chen, thank you
for always listening to my incessant rants and offering moral support. To Hanfei
Mao, thank you for your companionship over the years. To Annabel Huang, I am
grateful for the time spent with you in my last year, which made it the most
wonderful one among all these years. There are many more who I owe my gratitude
to: Chong and Nannan Wang, Sarah Xie, Tim Zhao, Fanlong Zeng, Luchang Jin, Yihan
Zhang, Daniel Fleischer, Junpu Wang, Hantao Yin \dots

Finally I would like to thank my parents. You have
been unconditionally supportive to whatever choices I make in life, whatever
dreams I pursue, and however far they take me from home. Without your support and
encouragement I would not have been able to make this far. This thesis is
dedicated to you.

\newpage \vspace*{8cm}
\pdfbookmark{Dedication}{dedication} % Sets a PDF bookmark for the dedication
\begin{center}
\large To the my parents, Bei Lin and Zhiping Chen
\end{center}
% Local Variables:
% TeX-master: "../thesis"
% End:
